Mathematics of Reinforcement Learning with Applications to Quantitative Finance

### 2022-08-17 papers/paper_2-col.tex
write up paper for this research

\documentclass[11pt]{article}
\usepackage{amsmath, amsfonts, bm, listings, enumitem, geometry, titlesec,
array, makecell, caption, graphicx, float, algorithm, algpseudocode, siunitx}
\usepackage[table]{xcolor}
\usepackage{tikz}
\usetikzlibrary{shapes.geometric, arrows.meta}
\usepackage{hyperref}
\geometry{margin=0.8in}
\graphicspath{ {../data/} }

%\renewcommand{\familydefault}{\sfdefault}
%\sisetup{output-exponent-marker=\ensuremath{\mathrm{e}}}
\tikzstyle{process} = [rectangle, minimum width=2cm, minimum height=1.3cm, 
text centered, draw=black]
\tikzstyle{arrow} = [thick,->,>={Stealth[scale=1.1]}]

\title{Reinforcement Learning Report}
\author{Yuanhao JIANG}
\date{}

\begin{document}
\maketitle
\tableofcontents

@@

\subsection{Policy Gradient Methods}
When the state space is arbitrarily large, we cannot find an optimal policy or
the optimal value function given the limited resources and time. Instead of 
keeping track of the value and the action(s) to select for each state (tabular 
methods), we can use parameterized functions, for example, artificial neural 
networks (ANNs), to approximate the value function and the policy function. 
\vspace{3mm}\newline Previously methods learned the values of actions and then 
select actions based on their estimated action values (q function). Now we 
consider methods that instead learn a parameterized policy that select actions 
without consulting a value function. A value function may still be used to 
learn the policy parameter, but is not required for action selection.
\vspace{3mm}\newline We use \(\bm{\theta}\) and \(\bm{w}\) to denotes the 
policy's parameter vector and value function's weight vector, respectively.
Then we have our policy \(\pi\left(a\;\middle|\;s, \bm{\theta}\right)=
\text{Pr}\left\{A_t=a\;\middle|\;S_t=s, \bm{\theta_t}=\bm{\theta}\right\}\) and
value function \(v_\pi(s, \bm{w})\). We also define the scalar performance 
measure (objective) to be 
\[J(\bm{\theta})=\begin{cases}
    \ v_{\pi_{\bm{\theta}}}(s_0); &\text{episodic case}\\
    \ \lim_{t \to \infty} \mathbb{E}
    \left[R_t\;\middle|\;S_0,A_{0:t-1}\sim\pi\right];      
                                  & \text{continuing case}
\end{cases}\]
Policy gradient methods seeks to maximize this performance measure, so their 
updates approximate gradient ascent in \(J\):
\[\bm{\theta}_{t+1}=\bm{\theta}_t+\alpha\widehat{\nabla J(\bm{\theta}_t)}\]
where \(\widehat{\nabla J(\bm{\theta}_t)}\) is a stochastic estimate whose
expection approximates the gradient of the performance measure w.r.t 
\(\bm{\theta}_t\).

@@

\subsection{Customer Features}
A customer' features is represented by an array (or vector), denoted by \(x\) 
with 16 entries with each generated randomly from a distribution, 
including normal distribution, binomial distribution, and so on. For example, 
a feature vector \(x\) is described as below:
\begin{table}[htpb]
    \centering
    \begin{tabular}{c|c|c|c|c|c|c|c}
        gender & age & car cost & miles & brand & random feature 0 & random 
        feature 1 & \dots\\
        \hline
        1.0 & 42.723 & 94032.096 & 30086.311 & 29.264 & 8.0 & 68.740 & \dots
    \end{tabular}
\end{table}

@@

\subsection{Resopnse}
The resopnse is generated by a generalized linear model, whose input is a 
vector of a customer features vector \(x\) appended with a price \(c\), i.e.
\(r=glm(x[0], \dots, x[15], c)\).
\vspace{3mm}\newline The GLM used in the project is fitted using a randomly
generated data set with each observation is a customer's features followed by 
a price, then followed by a resopnse e.g.
\begin{table}[htpb]
    \centering
    \begin{tabular}{c|c|c|c|c|c}
        & customer feature 0 & \dots & customer feature 15 & price & resopnse\\
        \hline
        1 & 1.0 & \dots & 2.0 & 1109.504 & 1.0\\
        \hline
        2 & 0.0 & \dots & 1.0 & 665.8 & 0.0\\
        \hline
        \dots & \dots & \dots & \dots & \dots & \dots
    \end{tabular}
\end{table}

\noindent And its link function is chose to be the logistic function.

@@

\subsection{Training Result}
The following graph shows the training result of the above algorithm (300
moving average reward vs iteration), with 600 iterations (5 episodes per 
iteration), \(T=300\), \(\gamma=0.99\), 
\(\alpha_\theta=\alpha_{\bm{w}}=\num{3e-4}\).
\begin{center}
    \includegraphics[scale=0.67]{a2c_out}
\end{center}
The graph shows that the moving average reward converges within only 50 
iterations, that is, 250 episode, 75000 parameter updates for each network, 
and stays quite stable afterwards.

@@

\subsection{Training result}
To make better comparison, the policy function of the two algorithms
is implemented the same way as we did in Actor-Critic algorithm, and the same 
is true for the value function used in PPO method. 
\vspace{3mm}\newline The following graph shows the training result of these 
three algorithms (300 moving average reward vs iteration), with 600 iterations
(5 episodes per iteration), \(T=300\), \(\gamma=0.99\), 
\(\alpha_\theta=\num{3e-4}\), \(\alpha_{\bm{w}}=\num{3e-10}\) (if applicable),
\(\epsilon=0.5\) (if applicable).
\begin{center}
    \includegraphics[scale=0.67]{out}
\end{center}
As is shown in the graph, the Actor-Critic algorithm converges the fastest,
within 50 iterations, followed by Reinforce algorithm, which converges within
about 250 iterations, despite a small drop of moving average around iteration 
300, which is quickly corrected by the algorithm within 50 iterations. PPO 
algorithm converges the slowest, with about 300 iterations. Despite of the 
speed of convergence, the moving average rewards of convergence for these 
three algorithms are nearly the same (around 400), this is as expected.

>>>>

% !TEX encoding = UTF-8 Unicode
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
%%
%%  This file is asmeconf-template.tex, a LaTeX template to format ASME Conference papers according to
%%  the requirements on ASME's conference web pages, and including hypertext support for the pdf.
%%
%%  This file is version 1.31 dated 2022/07/04
%%  
%%  As of version 1.11, this template defaults to ASME's newer conference guidelines first posted July 2019.
%% 			Those guidelines changed the requested author block formatting to be inline. 
%%			This LaTeX template continues to support the traditional grid format as a package option, [grid].
%%			Nomenclature now follows the abstract. Abstract text is set in italics.
%%
%%  Author: John H. Lienhard V
%%          Department of Mechanical Engineering
%%          Massachusetts Institute of Technology
%%          Cambridge, MA 02139-4307 USA
%%
%%  Class options include:
%%
%%          * Math options from M. Sharpe's newtxmath package: upright integrals [upint]; 
%%          *    [varvw] for a v and w that are better distinguished from Greek nu; and also 
%%          *    [subscriptcorrection, smallerops, varg, frenchmath, varbb, cmbraces, slantedGreek,...] 
%%			* 	 See newtx documentation for descriptions (at CTAN: http://ctan.org, v1.6 or higher is best).
%%
%%          * Options to: 	omit the ASME copyright footer [nofoot]; 
%%			*				use government employee copyright notice [govt];
%%			*				use government contractor copyright notice [contractor]
%%
%%          * An option to use the traditional grid arrangement of author names [grid].  
%%			*	 With this option, line breaks (\\) may be inserted into the address as needed. 
%%			*	 Author names that include commas should be enclosed in braces, e.g., {Joseph L. Smith, Jr.}.  
%%			*	 Authors may be grouped above a single affiliation using braces, e.g., {Henry Tudor, Catherine Parr}.
%%
%%          * An option to balance the heights of columns on the last page [balance]. 
%%          *    This option is NOT compatible with the [lineno] option.
%%
%%          * An option to include line numbers [lineno]. You must *run twice* for proper placement of the line numbers. 
%%			*	 The lineno package does not number titles, footnotes, captions, or tables.
%%          *    This option will disable balancing column height on final page if that option has been invoked.
%%          *    The lineno package won't always number the lines preceding displayed math in a paragraph because
%%          *    paragraph has not ended.  See that package's documentation for macros to address this problem, or
%%          *    just leave a blank line above the displayed equation while you are editing and then remove the 
%%          *    blank line and [lineno] option when you move to your final version.
%%
%%			* Options for PDF/A compliance. [pdf-a] will produce PDF/A-3u compliance with sRGB OutputIntent.
%%			*	 [pdfapart= 1 or 2 or 3] and [pdfaconformance= b or u] can enable levels 1b, 2b, 2u, and 3b.
%%			*    The most recent versions of LaTeX (2021 and later) are moving toward integrated support for pdf-a, 
%%			*    through \DeclareDocumentMetadata{..}. The asmeconf class supports these new features, which can 
%%			*	 replace the aforementioned class options. (An up-to-date LaTeX installation is required to use this.)
%%
%%          * Many options for calligraphic, script, and fraktur fonts from the mathalfa package; the
%%          *    example shown here is: [mathalfa=cal=boondoxo] to use a Boondox font for \mathcal.
%%          *    Some other options for cal are: dutchcal, zapfc, cm (default), euler,...
%%          *    frak (fraktur), bb (blackboard bold), scr (script) may also be chosen this way.
%%			*	 For details, refer to mathalfa documentation (at CTAN: http://ctan.org).
%%
%%          * Option to use superiors font from newtxtext for footnotes [nodefaultsups] and
%%          *    for slightly larger small capitals, [largesc], also from newtxtext.
%%
%%          * An option to allow hyphenation of the typewriter font [hyphenate], from inconsolata package.
%%          *    Hyphenation is normally suppressed for typewriter mode because it is often used for code.
%%			*	 To replace the default variable word spacing by monospacing, use the option [mono].
%%			*	 To get a zero without a slash, use [var0]
%%
%%          * Options (used by the babel package) to include passages in languages other than English (e.g., a translation 
%%			*    of the abstract). Languages are called as options, e.g. [french], [spanish], [greek], [russian], etc. 
%%			*    Language support is greatest when running LuaLaTeX with the [fontspec] option.
%%			*    See Appendix B for details.
%%
%%  The use of commands defined or modified by the asmeconf class is illustrated throughout this file. In particular, 
%%  ASME requires capitalized, sans-serif section headings, and as a result some care is needed when using macros 
%%  in section headings, as also illustrated below.
%%
%%  Use an up-to-date and complete installation of LaTeX, such as TeX Live 2020 or later. Errors may occur with old set-ups.
%%
 %=========================================================
%% 
%% LICENSE: 
%%
%% Copyright (c) 2022 John H. Lienhard
%%
%% Offered under the MIT license: https://ctan.org/license/mit 
%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% Class options are described above. Change these options as desired. 
%%		If you are not using the language options, remove them (together with Appendices B and C)
%%	 	Remove the [colorlinks] option before *final* submission to ASME, to get black text for printing,
%%		but keep that option for other uses.

\documentclass[balance,upint,subscriptcorrection,varvw,mathalfa=cal=boondoxo,pdf-a,colorlinks,nofoot]{asmeconf}
\usepackage{tikz}
\usetikzlibrary{shapes.geometric, arrows.meta}
\tikzstyle{process} = [rectangle, minimum width=2cm, minimum height=1.3cm, text centered, draw=black]
\tikzstyle{arrow} = [thick,->,>={Stealth[scale=1.1]}]
\usepackage{algorithm, algpseudocode}
\usepackage{siunitx}
\graphicspath{ {../data/} }
%%%%%  pdf metadata  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\hypersetup{%
	pdfauthor={Yuanhao JIANG},
	pdftitle={Mathematics of Reinforcement Learning with Applications to Quantitative Finance},
	pdfkeywords={Reinforcement Learning,  Artificial Neural Networks},
	pdfsubject={Describe Mathematics of Reinforcement Learning with Applications to Quantitative Finance},
    pdfurl={https://github.com/Yuanhao-JIANG/RL-in-QF},
%	pdflicenseurl={https://ctan.org/pkg/asmeconf},
}

\begin{document}

% Change these fields to the right content for your conference.
% You can comment these out if for some reason you don't want a header.
% Use title case (first letters capitalized), not all capitals

\ConfName{Proceedings of the ASME 2022\linebreak International Mechanical Engineering Congress and Exposition}
\ConfAcronym{IMECE2022}
\ConfDate{October 30--November 3, 2022} % update 
\ConfCity{Columbus, OH} % update 
\PaperNo{IMECE2022-XXXX}

% Units of measure (e.g., cm) and other specialty lowercase terms in the title should be 
%   enclosed in \NoCaseChange{...} to maintain lower case type
%   LaTeX will automatically set the rest of the title in all capital letters.

\title{Mathematics of Reinforcement Learning with Applications to Quantitative Finance}
%\title{Place Title Here: Place Subtitle After Colon}

%   Put author names into the order you want. Use the same order for affiliations.
%   \affil{#} tags the author's affiliation to the address in \SetAffiliation{#}.
%   No space between last name and \affil{#}, separate names with commas.
%
%	For a sole author or a single affiliation for all authors, {#} may be left empty, as \affil{} and \SetAffiliation{} (but not with [grid] option!)
%
%   \CorrespondingAuthor{email} follows that author's affiliation, no spaces.  
%   If multiple corresponding authors, put both email addresses in the same command and place after both authors.
%
%   \JointFirstAuthor, if applicable, follows the affiliation of the relevant authors, no spaces.

\SetAuthors{%
	Yuanhao JIANG\affil{1}\JointFirstAuthor\CorrespondingAuthor{}, 
	Luis Hern\'andez\affil{2}\JointFirstAuthor, 
	Catherine~Parr\affil{3}\CorrespondingAuthor{s2132254@ed.ac.uk, kate@thepalace.gov}
	}

\SetAffiliation{1}{The University of Edinburgh, Edinburgh, UK}
\SetAffiliation{2}{Institution or Company Name, City, State}
\SetAffiliation{3}{Hampton Court Palace, Richmond, England}
%	Note: Luis and Maria are not real people.  Henry and Catherine have been dead for >450 years.

%	To switch from inline author names to gridded names, use the [grid] option.

\maketitle

%%% Use this footnote for tracking various versions of your draft. Change text to suit your own needs. 
%%% \date{..} calls the same command. 
\versionfootnote{Documentation for \texttt{asmeconf.cls}: Version~\versionno, \today.}% <=== Delete before final submission.

%%% Change these to your keywords.  Keywords are automatically printed at the end of the abstract.
%%% This command MUST COME BEFORE the end of the abstract.
%%% If you don't want keywords, leave the argument of \keywords{} empty (or use the abstract* environment)

\keywords{Reinforcement Learning, Artificial Neural Networks}

%%%%%  End of fields to be completed. Now write your paper. %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%  ABSTRACT  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%
%% Abstract should be 200 words or less
\begin{abstract}

This paper is an example of and a  {\upshape\LaTeX} template for typesetting ASME conference papers using the {\upshape\texttt{asmeconf}} class. This  {\upshape\LaTeX} template follows ASME guidelines for margins, fonts, headings, captions, and reference formats as of 2022. The class is intended to be used with the {\upshape\texttt{asmeconf.bst} \hologo{BibTeX}} style for reference formatting, which is part of this distribution. The template produces pdfs that contain hyperlinks, bookmarks, and metadata; and references can include the DOI and URL fields. Links may be colored, for online use, or black, for publication. The class enables inline author names, following ASME's current style, but can also produce the traditional grid style. Options include line numbering, final column balancing, various math options, government copyright, and archivability (PDF/A). In addition, section headers may contain mathematics, references, citations, and footnotes. The class is compatible with {\upshape\hologo{pdfLaTeX}} or {\upshape\hologo{LuaLaTeX}}.
\end{abstract}

%%%%%%%%%  BODY OF PAPER %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\subsection{Policy Gradient Methods}
When the state space is arbitrarily large, we cannot find an optimal policy or
the optimal value function given the limited resources and time. Instead of
keeping track of the value and the action(s) to select for each state (tabular
methods), we can use parameterized functions, for example, artificial neural
networks (ANNs), to approximate the value function and the policy function.
\vspace{3mm}\newline Previously methods learned the values of actions and then
select actions based on their estimated action values (q function). Now we
consider methods that instead learn a parameterized policy that select actions
without consulting a value function. A value function may still be used to
learn the policy parameter, but is not required for action selection.
\vspace{3mm}\newline We use \(\bm{\theta}\) and \(\bm{w}\) to denotes the
policy's parameter vector and value function's weight vector, respectively.
Then we have our policy \(\pi\left(a\;\middle|\;s, \bm{\theta}\right)
=\text{Pr}\left\{A_t=a\;\middle|\;S_t=s, \bm{\theta}_t=\bm{\theta}\right\}\) and
value function \(v_\pi(s, \bm{w})\). We also define the scalar performance
measure (objective) to be
\[J(\bm{\theta})=\begin{cases}
    \ v_{\pi_{\bm{\theta}}}(s_0); &\text{episodic case}\\
    \ \lim_{t \to \infty} \mathbb{E}
    \left[R_t\;\middle|\;S_0,A_{0:t-1}\sim\pi\right];
                                  & \text{continuing case}
\end{cases}\]
Policy gradient methods seeks to maximize this performance measure, so their
updates approximate gradient ascent in \(J\):
\[\bm{\theta}_{t+1}=\bm{\theta}_t+\alpha\widehat{\nabla J(\bm{\theta}_t)}\]
where \(\widehat{\nabla J(\bm{\theta}_t)}\) is a stochastic estimate whose
expection approximates the gradient of the performance measure w.r.t
\(\bm{\theta}_t\).

@@

\subsection{Customer Features}
A customer' features is represented by an array (or vector), denoted by \(x\) 
with 16 entries with each generated randomly from a distribution, 
including normal distribution, binomial distribution, and so on. For example, 
a feature vector \(x\) is described as Table \ref{tab:1}
\begin{table*}
    \caption[Table]{Customer Feature Vector Description\label{tab:1}}
    \centering
    \begin{tabular}{c|c|c|c|c|c|c|c}
        gender & age & car cost & miles & brand & random feature 0 & random 
        feature 1 & \dots\\
        \hline
        1.0 & 42.723 & 94032.096 & 30086.311 & 29.264 & 8.0 & 68.740 & \dots
    \end{tabular}
\end{table*}

@@

\subsection{Resopnse}
The resopnse is generated by a generalized linear model, whose input is a 
vector of a customer features vector \(x\) appended with a price \(c\), i.e.
\(r=glm(x[0], \dots, x[15], c)\).
\vspace{3mm}\newline The GLM used in the project is fitted using a randomly
generated data set with each observation is a customer's features followed by 
a price, then followed by a resopnse, as illustrated by Table \ref{tab:2}.
\begin{table*}
    \caption[Table]{Customer Feature Vector Description\label{tab:2}}
    \centering
    \begin{tabular}{c|c|c|c|c|c}
        & customer feature 0 & \dots & customer feature 15 & price & resopnse\\
        \hline
        1 & 1.0 & \dots & 2.0 & 1109.504 & 1.0\\
        \hline
        2 & 0.0 & \dots & 1.0 & 665.8 & 0.0\\
        \hline
        \dots & \dots & \dots & \dots & \dots & \dots
    \end{tabular}
\end{table*}

\vspace{3mm}\noindent And the link function for the GLM is chosen to be the logistic function.

@@

\subsection{Training Result}
Figure \ref{fig:1} shows the training result of the above algorithm (300
moving average reward vs iteration), with 600 iterations (5 episodes per
iteration), \(T=300\), \(\gamma=0.99\),
\(\alpha_\theta=\alpha_{\bm{w}}=\num{3e-4}\).
\begin{figure*}
    \centering\includegraphics[scale=0.67]{a2c_out}
    \caption{Training Result for Algorithm a2c\label{fig:1}}
\end{figure*}
The graph shows that the moving average reward converges within only 50
iterations, that is, 250 episode, 75000 parameter updates for each network,
and stays quite stable afterwards.

@@

\subsection{Training result}
To make better comparison, the policy function of the two algorithms
is implemented the same way as we did in Actor-Critic algorithm, and the same
is true for the value function used in PPO method.
\vspace{3mm}\newline Figure \ref{fig:2} shows the training result of these
three algorithms (300 moving average reward vs iteration), with 600 iterations
(5 episodes per iteration), \(T=300\), \(\gamma=0.99\),
\(\alpha_\theta=\num{3e-4}\), \(\alpha_{\bm{w}}=\num{3e-10}\) (if applicable),
\(\epsilon=0.5\) (if applicable).
\begin{figure*}
    \centering\includegraphics[scale=0.67]{out}
    \caption{Training Result Comparison for Three Algorithm\label{fig:2}}
\end{figure*}
As is shown in Figure \ref{fig:2}, the Actor-Critic algorithm converges the fastest,
within 50 iterations, followed by Reinforce algorithm, which converges within
about 250 iterations, despite a small drop of moving average around iteration
300, which is quickly corrected by the algorithm within 50 iterations. PPO
algorithm converges the slowest, with about 300 iterations. Despite of the
speed of convergence, the moving average rewards of convergence for these
three algorithms are nearly the same (around 400), this is as expected.

\section{Conclusion}
Provide a brief conclusion (3-4 lines).

\section*{Acknowledgments}
Place any acknowledgments here.

%%%  REFERENCES  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%
%% Put your references into your .bib file in the usual way. Run latex once, bibtex once, then latex twice.
%% The asmeconf.bst style allows @inproceedings and @proceedings to include: 
%%		venue = {Location of Conference}, 
%%		eventdate = {Month, days},

\nocite{*}%% <=== Delete this line unless you want to typeset the entire contents of your .bib file!

\bibliographystyle{asmeconf}  %% .bst file following ASME conference format. Do not change.
\bibliography{paper_2-col}

%%%  APPENDICES  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\appendix

%% Note that appendices will be "numbered" A, B, C, ... etc. Use \section, not \section*
%% Equations will be numbered sequentially following those in the paper. Do not reset the equation counter.

%% Here we use the optional argument to control the pdf bookmark and prevent errors.
\section[The Vector Product A\times B]{The vector product $\vec{A}\times\vec{B}$}\label{appendix:a}

This brief illustration of an appendix shows the numbering of the appendix and equations. Equations are numbered
consecutively, following those in the paper. Consider $\rho \neq \textrm{fn}(p)$:
\begin{align}
\frac{d\Gamma}{dt} &{}= \frac{d}{dt} \int_{\mathcal{C}} \mathbf{u} \cdot d\mathbf{r}\\
				   &{}= \int_{\mathcal{C}} \frac{D\mathbf{u}}{Dt} \cdot d\mathbf{r} + \underbrace{\int_{\mathcal{C}} \mathbf{u}\cdot d\biggl( \frac{d\mathbf{r}}{dt}\Biggr)}_{=\, 0} \\[-2pt]
                   &{}= \iint_{\mathcal{S}} \nabla \times \frac{D\mathbf{u}}{Dt}  \cdot d\mathbf{A}\\
                   &{}= \iint_{\mathcal{S}}  \nabla p \times \nabla \left( \frac{1}{\rho}\right) \cdot d\mathbf{A}
\end{align}
